---
title: "PowerPuffGenes Report"
authors: "FN, CV, LP"
date: "2025-04-01"
output:
  html_document:
    toc: true
  pdf_document:
    toc: true
editor_options:
  chunk_output_type: console
---

# Introduction
Gene expression changes in response to stimuli can be broadly categorized as transient or prolonged. Transient changes are short-lived and may reflect immediate cellular responses such as signaling events or stress adaptation. In contrast, prolonged changes suggest stable regulatory shifts that may contribute to long-term cellular adaptations. Understanding these patterns can provide insight into gene regulatory mechanisms and cellular dynamics.

# Goals
- To classify genes exhibiting transient versus prolonged changes in expression after doxycycline (dox) induction.

# Gene Classification: Transient vs Prolonged

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE, echo = TRUE)
# library imports
library(IRanges)
library(dplyr)
library(tidyr)
library(tibble)
library(readr)
library(ggplot2)
library(purrr)
library(magrittr)
library(pheatmap)
library(textshape)
library(Rcpp)
library(matrixStats)
library(rtracklayer)
library(tidyverse)
library(DESeq2)
library(broom)
library(reshape)
```

## Load data
```{r loading our RNAseq analysis data COUNTS & TPM}
RESULT_PATH <- "../result_human"
COLUMN_PREFIX <- "gfp"
P_VALUE_FILTER <- 0.01
load(file.path(RESULT_PATH, "DESeq2_results.RData"))
load(file.path(RESULT_PATH, "TPM_results.RData"))
```

## Data curation
```{r assign group of genes to transient or prolonged}
# Filter rows where the comparison column contains "vs_0"
filtered_vs_zero <- filtered_results[grep("vs_0", filtered_results$comparison), ]

# Identify genes with significant changes at multiple time points (prolonged changes)
# Group by gene_id and filter genes that appear in more than one time point
prolonged_res <- filtered_vs_zero %>% group_by(gene_id) %>% filter(n() > 1)

# Extract gene names for prolonged genes specifically at time points containing "96_vs"
prolonged_genes <- prolonged_res[grep("96_vs", prolonged_res$comparison),
                                 which(names(prolonged_res) == "gene_name")]

# Merge the prolonged genes with the original filtered data to get detailed results
prolonged_res_results <- merge(filtered_vs_zero, prolonged_genes)

# Group prolonged genes by the specific time points where changes occur

# Identify genes with changes at time point "12_vs"
change_at_12 <- prolonged_res_results[grep("12_vs", prolonged_res_results$comparison),
                                      which(names(prolonged_res_results) == "gene_name")]

# Identify genes with changes at time point "24_vs"
change_at_24 <- prolonged_res_results[grep("24_vs", prolonged_res_results$comparison),
                                      which(names(prolonged_res_results) == "gene_name")]

# Identify genes with changes at time point "48_vs"
change_at_48 <- prolonged_res_results[grep("48_vs", prolonged_res_results$comparison),
                                      which(names(prolonged_res_results) == "gene_name")]

# Find genes that show consistent changes from 12 hours to the end (12, 24, and 48 hours)
change_at_12_to_end <- Reduce(intersect, list(change_at_12, change_at_24, change_at_48))

# Find genes that show consistent changes from 24 hours to the end (24 and 48 hours),
# excluding those already in the 12-to-end group
change_at_24_to_end <- intersect(change_at_24, change_at_48)
change_at_24_to_end <- change_at_24_to_end[!(change_at_24_to_end %in% change_at_12_to_end)]

# Find genes that show changes only at 48 hours, excluding those in previous groups
change_at_48_to_end <- change_at_48[!(change_at_48 %in% change_at_12_to_end) & 
                                    !(change_at_48 %in% change_at_24_to_end)]

# Identify genes with significant changes at a single time point (transient changes)
transient_genes <- filtered_vs_zero[!(filtered_vs_zero$gene_name %in% prolonged_genes$gene_name),
                                    which(names(filtered_vs_zero) == "gene_name")]
# Note: Transient genes are those that exhibit a spike in expression only at one specific time point.
# Save the identified gene groups into an RData file for later use
save(change_at_12_to_end, change_at_24_to_end, change_at_48_to_end, transient_genes,
     file = file.path(RESULT_PATH, "gene_groups.RData"))

# Create a data frame of genes with prolonged changes in expression, including the time point where changes start
prolonged_genes_df <- data.frame(
  gene_name = c(change_at_12_to_end, change_at_24_to_end, change_at_48_to_end),  # Combine all groups
  change_from = c(rep("12", length(change_at_12_to_end)),  # Label genes starting from 12 hours
                  rep("24", length(change_at_24_to_end)),  # Label genes starting from 24 hours
                  rep("48", length(change_at_48_to_end)))  # Label genes starting from 48 hours
)

# Filtering by tpm
# Extract unique gene IDs from the filtered results and create a data frame
sig_gene <- as.data.frame(unique(filtered_results$gene_id))  # Get unique gene IDs
names(sig_gene) <- "gene_id"  # Rename the column to "gene_id"

# Subset the data frame containing TPM values
tpm_filtered$gene_id <- rownames(tpm_filtered)
rownames(tpm_filtered) <- NULL
sig_gene_tpm_all <- inner_join(tpm_filtered, sig_gene, by = "gene_id")
sig_gene_tpm_all <- merge(sig_gene_tpm_all, g2s)  # Merge with gene-to-symbol mapping (g2s)

# Prepare TPM data for plotting genes of interest by merging with genes of interest
selected_tpm <- merge(sig_gene_tpm_all, prolonged_genes_df)

# Dynamically generate column names for 0h and 96h triplicates
t0_cols <- paste0(COLUMN_PREFIX, "_0_", 1:3)
t96_cols <- paste0(COLUMN_PREFIX, "_96_", 1:3)

# Function to perform t-test for one gene
calculate_stats <- function(gene_row) {
  tpm_0 <- as.numeric(gene_row[t0_cols])
  mean_0 = mean(tpm_0)
  tpm_96 <- as.numeric(gene_row[t96_cols])
  mean_96 = mean(tpm_96)
  log2FC = log2(mean(tpm_96) + 1e-3) - log2(mean(tpm_0) + 1e-3)
  t_result <- t.test(tpm_0, tpm_96, paired = FALSE)
  return(data.frame( pvalue = t_result$p.value,
                     tpm_log2fc = log2FC)
         )
}

# Apply function to each row/gene
stat_df <- do.call(rbind, apply(selected_tpm, 1, calculate_stats))
selected_tpm <- cbind(selected_tpm, stat_df)

# Adjust p-values for multiple testing (Benjamini-Hochberg)
selected_tpm$adjusted_pvalue <- p.adjust(selected_tpm$pvalue, method = "BH")

# Get significant results
significant_genes <- selected_tpm[selected_tpm$adjusted_pvalue < P_VALUE_FILTER, ]
genes_of_interest <- significant_genes %>% select(gene_name, change_from, tpm_log2fc)
genes_of_interest <- genes_of_interest[order(genes_of_interest$change_from),] 

# Save to csv file
write_csv(genes_of_interest, file.path(RESULT_PATH, "prolonged_gene_list.csv"))
```

## Data visualization
### LFC plot
```{r lfc versus time plot}
# Merge the main results with the genes of interest to extract relevant data for plotting
res_df <- merge(results, genes_of_interest)

# Filter comparisons against the baseline ("_vs_0") for time 0 data
df_time0 <- res_df %>% filter(grepl("_vs_0$", comparison))

# Extract the time (in hours) from the comparison column for time 0 data
df_time0 <- df_time0 %>%
  mutate(Time = as.numeric(gsub("time_([0-9]+)_vs_0", "\\1", comparison)))

# Filter out intermediate comparisons (not against baseline)
df_intermediate <- res_df %>% filter(!grepl("_vs_0$", comparison))

# Extract time points (Time1 and Time2) from the comparison column for intermediate data
df_intermediate <- df_intermediate %>%
  mutate(Time1 = as.numeric(gsub("time_([0-9]+)_vs_([0-9]+)", "\\1", comparison)),  # Start time
         Time2 = as.numeric(gsub("time_([0-9]+)_vs_([0-9]+)", "\\2", comparison)))  # End time

# Assign an offset index to stagger significance markers for each gene
df_intermediate <- df_intermediate %>%
  group_by(gene_name) %>%
  mutate(Offset_Index = row_number() - 1) %>%  # Create an index for each row
  ungroup()

# Calculate y-axis limits for each gene to ensure proper scaling and padding in plots
y_axis_limits <- res_df %>%
  group_by(gene_name) %>%
  summarize(
    min_log2FC = min(log2FoldChange, na.rm = TRUE),  # Minimum log2FoldChange value
    max_log2FC = max(log2FoldChange, na.rm = TRUE),  # Maximum log2FoldChange value
    y_axis_limit = max_log2FC + (max_log2FC - min_log2FC) * 0.05  # Add padding (5%)
  )

# Add y-axis limits and calculate positions for significance markers in intermediate data
df_intermediate <- df_intermediate %>%
  left_join(y_axis_limits, by = "gene_name") %>%  # Join y-axis limits with intermediate data
  mutate(
    Y_Pos = y_axis_limit + (Offset_Index * 0.3)  # Stagger significance markers above the max limit
  )

# Define significance levels based on p-values and create labels for markers (***, **, *)
df_intermediate <- df_intermediate %>%
  mutate(Significance = case_when(
    pvalue < 0.001 ~ "***",   # Highly significant
    pvalue < 0.01 ~ "**",     # Significant
    pvalue < 0.05 ~ "*",      # Marginally significant
    TRUE ~ ""                 # Not significant
  ))

# Create a ggplot visualization with facets for each gene
ggplot() +
  
  # Plot log2FoldChange values for baseline comparisons (time vs. baseline)
  geom_point(data = df_time0, aes(x = Time, y = log2FoldChange, color = padj < 0.05), size = 1.5) +
  
  # Add lines connecting points for baseline comparisons within each gene
  geom_line(data = df_time0, aes(x = Time, y = log2FoldChange, group = gene_name)) +
  
  # Add horizontal segments for p-value markers in intermediate comparisons (significant only)
  geom_segment(data = df_intermediate %>% filter(Significance != ""), 
               aes(x = Time1, xend = Time2, 
                   y = Y_Pos, yend = Y_Pos), 
               color = "black", size = 0.5) +

  # Add significance labels (***, **, *) at the midpoint of each segment
  geom_text(data = df_intermediate %>% filter(Significance != ""), 
            aes(x = (Time1 + Time2) / 2, y = Y_Pos + 0.025, label = Significance), 
            size = 4, color = "black") +

  # Facet wrap: Separate plots for each gene with independent scales ("free")
  facet_wrap(~gene_name, scales = "free") +

  # Customize x-axis ticks to display specific time points of interest (12, 24, etc.)
  scale_x_continuous(breaks = c(12, 24, 48, 96), labels = c("12", "24", "48", "96")) +
  
  # Add titles and axis labels with clear formatting
  labs(title = "Genes regulated by dox over time",
       caption = "*: p < 0.05\n**: p < 0.01\n***: p < 0.001",
       x = "Time after dox induction (hours)",
       y = "log2fold change",
       color = "Significant (padj < 0.05)\nto -dox control") +
  
   # Apply a minimal theme with customizations for clarity and aesthetics
   theme_minimal() +
   theme(
     panel.grid.major = element_blank(), 
     panel.grid.minor = element_blank(),
     panel.background = element_rect(fill = 'white', colour = 'white'),
     panel.border = element_rect(color = "black", fill = NA, linewidth = 1)
   )

# Save the plot as a high-resolution PNG file in the results directory
ggsave(file.path(RESULT_PATH, "lfc_prolonged_genes.png"), width = 20, height = 12, dpi = 500)
```

### TPM plot
```{r TPM for genes of interest}
# Extract unique gene IDs from the filtered results and create a data frame
sig_gene <- as.data.frame(unique(filtered_results$gene_id))  # Get unique gene IDs
names(sig_gene) <- "gene_id"  # Rename the column to "gene_id"

# Subset the data frame containing average and standard deviation (SD) TPM values
sig_gene_tpm <- inner_join(avg_and_sd_values, sig_gene, by = "gene_id")  # Join with average/SD values
sig_gene_tpm <- merge(sig_gene_tpm, g2s)  # Merge with gene-to-symbol mapping (g2s)

# Prepare TPM data for plotting genes of interest by merging with genes of interest
df_plot_genes <- merge(sig_gene_tpm, genes_of_interest)

# Extract column names corresponding to average TPM values
time_avg_values <- names(df_plot_genes)[grep("avg", names(df_plot_genes))]

# Reshape the data to long format for average TPM values
df_plot_genes_avg <- melt(df_plot_genes, measure.vars = time_avg_values, value_name = "gene_id")
df_plot_genes_avg$Time <- do.call(rbind, stringr::str_split(df_plot_genes_avg$variable, '_avg'))[,1]  # Extract time
df_plot_genes_avg$Time <- as.numeric(df_plot_genes_avg$Time)  # Convert time to numeric
colnames(df_plot_genes_avg)[which(colnames(df_plot_genes_avg) == 'value')] <- 'avg'  # Rename value column to 'avg'
df_plot_genes_avg <- df_plot_genes_avg %>% select("gene_name", "Time", "avg")  # Select relevant columns

# Extract column names corresponding to SD TPM values
time_sd_values <- names(df_plot_genes)[grep("sd", names(df_plot_genes))]

# Reshape the data to long format for SD TPM values
df_plot_genes_sd <- melt(df_plot_genes, measure.vars = time_sd_values, value_name = "gene_id")
df_plot_genes_sd$Time <- do.call(rbind, stringr::str_split(df_plot_genes_sd$variable, '_sd'))[,1]  # Extract time
df_plot_genes_sd$Time <- as.numeric(df_plot_genes_sd$Time)  # Convert time to numeric
colnames(df_plot_genes_sd)[which(colnames(df_plot_genes_sd) == 'value')] <- 'sd'  # Rename value column to 'sd'
df_plot_genes_sd <- df_plot_genes_sd %>% select("gene_name", "Time", "sd")  # Select relevant columns

# Merge average and SD data frames on gene_name and Time
df_plot_genes_avg_sd <- merge(df_plot_genes_avg, df_plot_genes_sd, by = c("gene_name", "Time"))

# Create a ggplot visualization for TPM values across time points
ggplot(df_plot_genes_avg_sd, aes(x = Time, y = avg)) +
  geom_line() +  # Add lines connecting average TPM values for each gene over time
  geom_point(size=1) +  # Add points at each time point
  geom_errorbar(aes(ymin = avg-sd, ymax = avg+sd), width = 3, size = 0.5) +  # Add error bars using SD
  labs(x = 'Time after dox induction (hours)', y = 'Average TPM') +  # Label axes
  ggtitle('Average TPM across Time Points') +  # Add a title to the plot
  facet_wrap(~ gene_name, scales = "free") +  # Create separate plots for each gene with independent scales
  
  # Customize x-axis ticks to show specific time points of interest
  scale_x_continuous(breaks = c(0, 12, 24, 48, 96), labels = c("0", "12", "24", "48", "96")) +
  
  theme_minimal() +  # Use a minimal theme for clean visuals
  theme(
    panel.grid.major = element_blank(), 
    panel.grid.minor = element_blank(),
    panel.background = element_rect(fill = 'white', colour = 'white'),
    panel.border = element_rect(color = "black", fill = NA, size = 1)
    )

# Save the plot as a high-resolution PNG file in the results directory
ggsave(file.path(RESULT_PATH, "tpm_prolonged_genes.png"), width = 20, height = 12, dpi = 500)
```


