---
title: "PowerPuffGenes Report"
authors: "FN, CV, LP"
date: "2025-04-16"
output:
  html_document:
    toc: true
  pdf_document:
    toc: true
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
source("../utils/ATAC_functions.R")
library(GenomicRanges)
library(IRanges)
library(dplyr)
library(tidyr)
library(tibble)
library(readr)
library(ggplot2)
library(purrr)
library(magrittr)
library(pheatmap)
library(textshape)
library(Rcpp)
library(DESeq2)
library(stringr)
```

#### Objective: To determine if chromatin accessibility changes due to dox exposure in human stem cells.
We have performed a time course series of experiements measuring chromating accessibility
(ATACseq peaks)upon exposure to dox. These are 0, 30, 60, 90, 120, 150 minutes.
Each time point has a replicate. The fastq sequencing files were processed by the
NF_CORE ATACseq pipeline v- 2.12 (https://nf-co.re/atacseq/2.1.2/).

We will be using the MACS2 output files from the NF_CORE Pipeline For example:
-broad.peak
-consensus.broad.peak
-consensus.feature.counts

#### Approach:

#### Parameters and paths definition
```{r define variables}
PEAK_PATH         <- "/scratch/Shares/rinnclass/MASTER_CLASS/DATA/mouse_atacseq"
RESULT_PATH       <- "../result_mouse"

```

##### (a) Loading in ATACseq peak files with custom function import_peaks (list of GRanges output)
```{r loading in peak files to list of GRanges}
# creating a file list also needed for import_peaks function to get sample name associated with file
fl <- list.files(PEAK_PATH, full.names = TRUE, pattern = ".broadPeak")

# running import_peaks
my_peaks <- import_peaks(PEAK_PATH)

# finding consensus peaks among replicates
my_consensus_peaks <- find_consensus_peaks(my_peaks)

print("here are the number of peaks for each sample")
print(num_consensus_peaks <- sapply(my_consensus_peaks, length) %>% as.data.frame)
```

##### Result - there are roughly 100-160K peaks in a given sample
##### (b) finding number of peaks common in all samples using find_common_peaks custom funciton
```{r finding common peaks in all samples, dependson="previous_chunk"}
# run find_common_peaks function
common_peaks <-  suppressWarnings(find_common_peaks(my_consensus_peaks))

```
Here are the number of peaks that overlap in all samples: `r length(common_peaks)`


##### (c) Peaks that are unique to dox and non-dox conditions
Using find_common_overlaps peaks that are specific to dox or non-dox
will be identified.

Non-dox samples
```{r non-dox atac peaks, dependson="previous_chunk"}

# common peaks in non-dox (0 time point)
# TODO: What does KO mean?
non_dox_samples <- names(my_consensus_peaks)[c(grep("_0", names(my_consensus_peaks)))]
non_dox_peaks <- my_consensus_peaks[non_dox_samples]
non_dox_common_peaks <- suppressWarnings(find_common_peaks(non_dox_peaks))
if (length(non_dox_samples) > 1) {
  non_dox_common_peaks <- suppressWarnings(find_common_peaks(non_dox_peaks))
} else {
  non_dox_common_peaks <- non_dox_peaks[[1]]
  }
print(c("This is how many peaks are common in non-dox:", length(non_dox_common_peaks)))
```

Dox samples
```{r dox samples, dependson="previous_chunk"}
# common peaks in dox time points (!not time 0)
dox_samples <- names(my_consensus_peaks)[!grepl("_0$", names(my_consensus_peaks))]
dox_peaks <- my_consensus_peaks[dox_samples]
dox_union_peaks <- suppressWarnings(find_union_peaks(dox_peaks))
dox_common_peaks <- suppressWarnings(find_common_peaks(dox_peaks))
# TODO: Should we use union_peaks or common_peaks?

print(c("This is how many unique peaks are in dox samplex",length(dox_union_peaks)))
print(c("This is how many peaks are common in dox",length(dox_common_peaks)))
```


Overlap of dox and non-dox (to get unique to each condition)
```{r dox vs non-dox atac peaks, dependson="previous_chunk"}
# Now overlap between dox and non-dox common peaks
dox_compare_list <- list(non_dox = non_dox_common_peaks, dox = dox_common_peaks)
dox_non_dox_ov <- suppressWarnings(find_common_peaks(dox_compare_list))
print(c("This is how many peaks are common in both non- and dox treatments", length(dox_non_dox_ov)))

# extracting peaks unique to each condition (dox non-dox)
# Peaks unique to non_dox
unique_to_non_dox <-suppressWarnings(find_my_peaks(dox_non_dox_ov, non_dox_common_peaks))
print(c("This is how many peaks are unique to non-dox condition", length(unique_to_non_dox)))

# Peaks unique to dox
unique_to_dox <- suppressWarnings(find_my_peaks(dox_non_dox_ov, dox_common_peaks))

print(c("This is how many peaks are unique to dox condition", length(unique_to_dox)))
```

#### Summary dox non-dox overlaps (common peaks)
##### Result: 37348 dox common peaks of the 53747 non-dox peaks overlapped (69.5%)
##### Result: 16399 peaks found in non-dox (30.5%) that were not found in dox common
##### Result: 1774 unique peaks in dox (1%) and not present in non-dox


Overlap of dox and non-dox (to get unique to each condition)
```{r dox vs non-dox atac peaks, dependson="previous_chunk"}
# Now overlap between dox and non-dox common peaks
dox_compare_list <- list(non_dox = non_dox_common_peaks, dox = dox_union_peaks)
dox_non_dox_ov <- suppressWarnings(find_common_peaks(dox_compare_list))
print(c("This is how many peaks are common in both non- and dox treatments", length(dox_non_dox_ov)))

# extracting peaks unique to each condition (dox non-dox)
# Peaks unique to non_dox
unique_to_non_dox <-suppressWarnings(find_my_peaks(dox_non_dox_ov, non_dox_common_peaks))
print(c("This is how many peaks are unique to non-dox condition", length(unique_to_non_dox)))

# Peaks unique to dox
unique_to_dox <- suppressWarnings(find_my_peaks(dox_non_dox_ov, dox_union_peaks))

print(c("This is how many peaks are unique to dox condition", length(unique_to_dox)))
```

#### Summary dox non-dox overlaps (union peaks)

##### Result: almost all peaks found in non-dox were also present in dox
##### Result: 53584 dox union peaks of the 53747 non-dox common peaks (99.7%)
##### Result: 163 peaks found in non-dox (0.3%) that were not found in dox union
##### Result: 112266 union peaks in dox (67.9%) and not present in non-dox
