---
title: "PowerPuffGenes Report"
authors: "FN, CV, LP"
date: "2025-04-01"
output:
  html_document:
    toc: true
  pdf_document:
    toc: true
editor_options:
  chunk_output_type: console
---

# Introduction
- Doxycycline (or Dox) is an antibiotic drug used in biological research for inducing gene expression. 
intro in dox, 
experiment 
For this analysis, we used wild type mouse (Mus musculus) stem cells at different time points after Dox treatment was applied. We used three replicates for each time point: 0, 12, 24, 48 and 96 hours after treatment.  

# Goals
- Identify specific genes that are significantly up or downregulated after Dox treatment at various time points. 
- General analysis of gene expression. Consider repeating the analysis from class but normalizing based on time 0.

# Methods
##Nextflow RNA Sequencing Pipeline
- For our analysis, we used Nextflow's RNAseq pipeline versions [3.14.0.][https://nf-co.re/rnaseq/3.14.0] and [3.18.0.][https://nf-co.re/rnaseq/3.18.0]  

- The inputs we used for running this pipeline: 
 - Raw fasta sample sequence files
 - Mouse reference genome assembled sequence (GRCm38/mm10)
 - Genome annotation 

- This pipeline consists of a series of steps that process our samples  to give an output raw read counts, and transcripts per million(tpm). Counts and tpm values were used to 

## Nextflow output
- May describe the results from the nextflow pipeline, some QC plots and the type of files produced.

##DESeq2 Analysis
- Explain in general the overall idea with RNA-seq analysis

# Data preprocessing for RNA-seq analysis with DESeq2

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE, echo = TRUE)
# library imports
library(IRanges)
library(dplyr)
library(tidyr)
library(tibble)
library(readr)
library(ggplot2)
library(purrr)
library(magrittr)
library(pheatmap)
library(textshape)
library(Rcpp)
library(matrixStats)
library(rtracklayer)
library(tidyverse)
library(DESeq2)
library(broom)
library(reshape)
```

### Parameters and paths definition
```{r define variables}
COUNTS_PATH      <- "../data/NF3-14_rep1/salmon.merged.gene_counts.tsv"
TPM_PATH         <- "../data/NF3-14_rep1/salmon.merged.gene_tpm.tsv"
FILTER_TRESHOLD  <- 0
P_VALUE_FILTER   <- 0.05
LOG2_FOLD_FILTER <- 1
```

### Counts loading and preprocesing
```{r load counts and get counts filtered}
# Read raw count data and make dictionary mapping gene id with gene name
counts_orig  <- read.table(COUNTS_PATH, header=TRUE, row.names=1)
g2s          <- data.frame(gene_id = rownames(counts_orig), 
                           gene_name = counts_orig[ , 1]) 
write.csv(g2s, "../results/g2s.csv")

# Remove gene name column from counts and convert counts to numerical rounded matrix
counts           <- counts_orig %>% select(-gene_name)
counts_matrix    <- as.matrix(counts) 
counts_matrix    <- round(counts_matrix)

# Filter genes with no counts for any of the samples
counts_filtered  <- counts_matrix[rowSums(counts_matrix) > FILTER_TRESHOLD, ]
```

### Metadata preparation
```{r make deseq_samples metadata file for DESeq2 }
# Make a column from the titles of the columns of the counts_matrix table
deseq_samples  <- data.frame(sample_id = colnames(counts))

# Get the time and replicate values from the sample names in deseq_samples
split_values     <- strsplit(deseq_samples$sample_id, "_")
time_values      <- sapply(split_values, function(x) x[[2]])
replicate_values <- sapply(split_values, function(x) x[[3]])

# Add time and replicate values as columns to deseq_samples and factor them
deseq_samples$time_point <- time_values
deseq_samples$replicate  <- replicate_values
deseq_samples$time_point <- factor(deseq_samples$time_point, levels=c("0","12","24","48","96"))
deseq_samples$replicate  <- factor(deseq_samples$replicate)

# Testing whether sample sheet and counts are arranged properly 
stopifnot(all(colnames(counts) == rownames(deseq_samples$sample_id)))
```

### DESeq execution
```{r run DESeq and obtain rlogs}
# Prepare DESeq dataset and run DESea2
dds <- DESeqDataSetFromMatrix(countData = counts_filtered,
                              colData = deseq_samples,
                              design = ~ time_point) 
dds <- DESeq(dds)
```

### DESeq time-point arrangement
```{r obtain results for all comparisons}
# Define the comparisons
comparison_list <- list(
  # Time points against baseline
  time_12_vs_0 = c("time_point", "12", "0"),
  time_24_vs_0 = c("time_point", "24", "0"),
  time_48_vs_0 = c("time_point", "48", "0"),
  time_96_vs_0 = c("time_point", "96", "0"),
  # Time series comparisons
  time_24_vs_12 = c("time_point", "24", "12"),
  time_48_vs_24 = c("time_point", "48", "24"),
  time_96_vs_48 = c("time_point", "96", "48")
)

# Create empty df to store results values (schema looks good)
results <- data.frame("gene_id" = character(),
                      "baseMean" = numeric(),
                      "log2FoldChange" = numeric(),
                      "lfcSE" = numeric(),
                      "stat" = numeric(),
                      "pvalue" = numeric(),
                      "padj" = numeric(),
                      "gene_name" = character(),
                      "comparison" = character())

# Loop through the defined contrasts
for(i in names(comparison_list)) {
  current_comparison <- comparison_list[[i]] # get comparison i
  res <- results(dds, contrast = current_comparison) # get results for time comparison i using 'contrast'
  if (is.null(res)) {
      warning(paste("Could not generate results for contrast:", contrast_name))
      next}
  # Temporary df to store the results for comparison i
  tmp_res_df <- res %>% 
    as.data.frame() %>%
    rownames_to_column("gene_id") %>%
    merge(g2s, by = "gene_id", all.x = TRUE) %>% 
    mutate(comparison = i) 
  # Add the temporary df to the main results df
  results <- bind_rows(results, tmp_res_df)
}
```

### Filtering and visualization
```{r filtering results}
# Filter based on p-value < 0.05 and log2 fold change > 1
filtered_results <- results %>%
  filter(padj < P_VALUE_FILTER, abs(log2FoldChange) > LOG2_FOLD_FILTER)

# Get all gene names that are significant (drop gene name repetitions)
filtered_genes <- as.data.frame(filtered_results$gene_name, collapse = "\n")
filtered_genes <- unique(filtered_genes)
colnames(filtered_genes)[1] <- "gene_name"

# Now let's write this out and do gene enrichment analysis
write.table(filtered_genes["gene_name"], row.names = FALSE, col.names = FALSE, "../results/filtered_genes.csv")

save(results, filtered_results, filtered_genes, deseq_samples, g2s, file = "../results/DESeq2_results.RData")
```

```{r filtered results distributions}
# Distribution of baseMean, lfcSE, p-values and fold change
#hist(filtered_results$baseMean, xlim = c(0, 10000), breaks = 500) # May remove this plot
#hist(filtered_results$lfcSE, xlim = c(0,5), breaks = 100) # May remove this histogram as well
hist(filtered_results$padj, xlim = c(0,P_VALUE_FILTER ), breaks = 50)
hist(filtered_results$log2FoldChange, xlim = c(-5, 5), breaks = 1000)

# T-test comparing baseMean expression of all genes vs filtered genes
basemean_all_genes <- median(results$baseMean)
basemean_fil_genes <- median(filtered_results$baseMean)
t.test(results$baseMean, filtered_results$baseMean)

# Intersect by gene_name filtered results table and raw counts
counts_filtered_genes <- counts_orig[filtered_results$gene_id,]
# TODO delete repetitive instances of genes
```

## TPM analysis
```{r, load tpm data and filter}
# Read raw tpm data and delete gene name column
tpm_orig  <- read.table(TPM_PATH, header=TRUE, row.names=1)
tpm       <- tpm_orig %>% select(-gene_name)

# Filter genes with no counts for any of the samples
tpm_filtered <- tpm[rowSums(tpm) > FILTER_TRESHOLD, ]
```

```{r calculate avg and sd for tpms}
# Loop to calculate avg and sd for replicates at a given time point
time_points <- c("0", "12", "24", "48", "96")
avg_and_sd_values <- list()
for (tp in time_points) {
  cols     <- grep(paste0("WT_", tp, "_"), colnames(tpm_filtered))
  avg      <- rowMeans(tpm_filtered[, cols])
  sd       <- apply(tpm_filtered[, cols], 1, sd)
  sd       <- data.frame(sd)
  combined <- cbind(avg, sd)
  avg_and_sd_values <- c(avg_and_sd_values, list(combined))
}

# Convert the list to a data frame and add column names for the respective timepoint
avg_and_sd_values           <- do.call(cbind,  avg_and_sd_values)
colnames(avg_and_sd_values) <- paste0(rep(time_points, each = 2), c("_avg", "_sd"))
avg_and_sd_values           <- as.data.frame(avg_and_sd_values) %>% rownames_to_column("gene_id") %>% merge(g2s)

save(tpm_orig, tpm_filtered, avg_and_sd_values, g2s, deseq_samples, file = "../results/tpm_results.RData")
```

```{r facet all timepoint TPM distributions}
# Make a distribution of TPM values in each time point - using facet-wrap
time_values <- names(avg_and_sd_values)[grep("avg", names(avg_and_sd_values))]
melt_tpm_df <- melt(avg_and_sd_values, measure.vars = time_values, value_name = "gene_id")
ggplot(melt_tpm_df, aes(x = value)) +
  geom_histogram(bins = 50) +
  facet_wrap(~ variable, scales = "free_x")+
  xlim(0,1000) +
  ylim(0,1000) +
  theme("paperwhite")
```























  